\texorpdfstring{\ZpL}{ZpL}: a p-adic precision package

### 2018-05-07 ZpL/ZpL.tex
More referee changes

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
\newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
\newcommand{\QpLC}{\text{\color{output} \rm \tt QpLC}\xspace}
\newcommand{\QpLF}{\text{\color{output} \rm \tt QpLF}\xspace}
\newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}

@@

\newcommand{\cIn}{{\color{input} \tt \phantom{Zp}In}:}
\newcommand{\cOut}{{\color{red} \tt \phantom{Z}Out}:}
\newcommand{\cZpCR}{{\color{output} \tt ZpCR}:}
\newcommand{\cZpFP}{{\color{output} \tt ZpFP}:}
\newcommand{\cZpLC}{{\color{output} \tt ZpLC}:}
\newcommand{\cZpLF}{{\color{output} \tt ZpLF}:}

@@

If provably correct results are desired, interval arithmetic provides
an alternative to floating point.  Instead of just tracking an approximation to the answer,
the package also tracks a radius within which the true result lies.
This method is commonly used for $p$-adic computations since
the ultrametric property of $p$-adic fields frequently keeps the radius small.
Computations remain fairly efficient with this approach, but numerical instability
can still lead to dramatic losses in precision (see \S \ref{sec:demo} for many examples).
Tracking the precision of multiple variables concurrently, the set of possible
true values associated to an inexact value takes the form of an ellipsoid
with axes parallel to the coordinate axes.

@@

In this paper, we describe an implementation of this idea in \sage \cite{sage}.
Rather than attaching a precision to each element, we store the precision of
many elements together by tracking a precision module\footnote{Throughout
the paper, the word ``module'' has its mathematical meaning: it denotes 
an additive group endowed with an external product by scalars in $\Zp$.} 
for the whole collection of variables.  As variables are created and 
destroyed, we update a matrix whose rows represent the vectors in the 
module.  Information about the precision of elements is extracted from 
the matrix as necessary.
Concretely, our package provides two new \sage 
parents\footnote{Parents in \sage are the analogues of container objects 
in computer science. Basically it corresponds to the implementation of a 
concrete mathematical structure, as the ring of $p$-adic integers.}, 
namely \ZpLC and \ZpLF (and their fraction fields \QpLC and \QpLF). The 
acronyms \LC and \LF stand for ``lattice-cap'' and ``lattice-float''
respectively (see \S \ref{sec:implementation} for details). Our package 
is merged in the standard distribution of \sage since version 8.2.

@@

\noindent
Again \ZpCR does not output the optimal precision when the computation
is split into several steps whereas \ZpL does.
Actually, these toy examples illustrate quite common situations which 
often occur during the execution of many algorithms. For this reason,
interval arithmetic often overestimates the losses of precision. 
Roughly speaking, the aim of our package is to ``fix this misfeature''.
In the next subsections, we present a bunch of examples showing the
benefit of \ZpL in various contexts.

@@

Many generic algorithms of linear algebra lead to quite important 
instability when they are used with $p$-adic numbers. In many cases, 
our package \ZpL rubs this instability without having to change the 
algorithm, nor the implementation.

@@

\noindent \textbf{Characteristic polynomials.}
%
Characteristic polynomials are notoriously hard to compute 
\cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}. We illustrate this 
with the following example (using the default algorithm of \sage for the 
computation of the characteristic polynomial, which is a division free 
algorithm in this setting) :

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
 & \verb?M.?{\color{method}\verb?determinant?}\verb?()? \\
\cZpCR
 & \verb?...010000010? \\
\cZpLC
 & \verb?...010000010? \\
\cIn
 & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
\cZpCR
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
\cZpLC
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
 & \verb?...010000010? \\
\end{tabular}}

@@

{\noindent \small
\begin{tabular}{@{}rl}
\cIn
% & {\color{ring}\verb?Q2?}\verb? = Qp(2, prec=50, print_mode="digits")? \\
 & \verb?R.<x,y,z> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, order=?{\color{string}\verb?'invlex'?}\verb?)? \\
 & \verb?F = [ ?{\color{ring}\verb?Q2?}\verb?(2,10)*x + ?{\color{ring}\verb?Q2?}\verb?(1,10)*z, ? \\
 & \verb?      ?{\color{ring}\verb?Q2?}\verb?(1,10)*x^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y^2 - ?{\color{ring}\verb?Q2?}\verb?(2,10)*z^2,? \\
 & \verb?      ?{\color{ring}\verb?Q2?}\verb?(4,10)*y^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y*z + ?{\color{ring}\verb?Q2?}\verb?(8,10)*z^2 ]? \\
\cIn
 & {\color{keyword}\verb?from?}\verb? sage.rings.polynomial.toy_buchberger\?\\
 & \verb?    ?{\color{keyword}\verb?import?}\verb? ?{\color{function}\verb?buchberger_improved?} \\
 & \verb?g = ?{\color{function}\verb?buchberger_improved?}\verb?(ideal(F))? \\
 & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
\cZpCR
 & \verb?[x^3, x*y + ...1100010*x^2,? \\
 & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
\cZpLC
 & \verb?[x^3, x*y + ...111100010*x^2,? \\
 & \verb? y^2 + ...1111111001*x^2, z + ...0000000010*x]? \\
\end{tabular}}

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
% & \verb?N = 4? \\
% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=50, print_mode="digits")? \\
% & {\color{ring}\verb?Q2?}\verb?=?{\color{ring}\verb?Z2?}\verb?.fraction_field()? \\
 & \verb?S.<t> = ?{\color{constructor}\verb?PowerSeriesRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, 16)? \\
 & \verb?h = 1 + t + t^3? \\
 & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S, prec=10)? \\
 & \verb?g = y.derivative() / h(y)? \\
 & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4); u[15]? \\
\cZpCR
 & \verb?...1101? \\
\cZpLC
 & \verb?...11011101? \\
\end{tabular}}

@@

Let us first examine the first construction [Computation].
With our conventions, if $t$ is the time just before the execution of 
the instruction we are interested in, the $\ttv_i$'s lie in $\calV_t$ 
while $\ttw$ does not. Moreover $\calV_{t+1} = \calV_t \sqcup \{\ttw\}$, 
so that $E_{t+1} = E_t \oplus \Qp e_\ttw$.
The mapping taking the values of variables at time $t$ to that at 
time $t{+}1$ is:
$$\begin{array}{rcl}
F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n)
\end{array}$$
where $x_i$ is the $\ttv_i$-th coordinate of the vector $\underline x$.
The Jacobian matrix of $F$ at $\underline x$ is easily computed; it 
is the block matrix $J_{\underline x}(F) = \big( \begin{matrix} I & L
\end{matrix} \big)$ where $I$ is the identity matrix of size $\text{Card}
\:\calV_t$ and $L$ is the column vector whose $\ttv$-th entry is
$\frac{\partial f}{\partial \ttv} (\underline x)$ if $\ttv$ is one of 
the $\ttv_i$'s and $0$ otherwise.
Therefore, the image of $H_t$ under $dF_{\underline x}$ is represented 
by the matrix $J_{\underline x}(F) \cdot M_t = \big( \begin{matrix} M_t & C
\end{matrix} \big)$ where $C$ is the column vector:
\begin{equation}
\label{eq:defC}
C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
\cdot C_i
\end{equation}
where $C_i$ is the column vector of $M_t$ corresponding to the
variable $\ttv_i$.
Observe that the matrix $J_{\underline x}(F) \cdot M_t$ is no longer a 
square matrix; it has one extra column. This reflects the fact that 
$\dim E_{t+1} = \dim E_t + 1$. Rephrasing this in a different language, 
the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
$E_{t+1}$ but is included in an hyperplane.

The package \ZpLC tackles this issue by introducing a cap: we do not 
work with $dF_{\underline x}(H_t)$ but instead define the lattice
$H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$
where $N_{t+1}$ is an integer, the so-called \emph{cap}.
Alternatively, one may introduce the map:
\begin{equation}
\label{eq:tildeF}
\begin{array}{rcl}
\tilde F: \quad E_t \oplus \Qp & \longrightarrow & E_{t+1} \smallskip \\
\underline x \oplus c & \mapsto & \underline x \:\oplus 
\big(f(x_1, \ldots, x_n) + c\big).
\end{array}
\end{equation}
The lattice $H_{t+1}$ is then the image of $H_t \oplus p^{N_{t+1}}
\Zp$ under $d\tilde F_{(\underline x, \star)}$ for any value of $\star$.
The choice of the 
cap is of course a sensitive question. \ZpLC proceeds as follows. When a 
ring is created, it comes with two constants (which can be specified by 
the user): a relative cap \textsc{relcap} and an absolute cap 
\textsc{abscap}. With these predefined values, the chosen cap is:
\begin{equation}
\label{eq:cap}
N_{t+1} = 
\min\big(\textsc{abscap},\, \textsc{relcap} + v_p(y)\big)
\end{equation}
with $y = f(x_1, \ldots, x_n)$.
In concrete terms, the lattice $H_{t+1}$ is represented by the block 
matrix:
$$\left(\begin{matrix}
M_t & C \smallskip \\ 0 & p^{N_{t+1}}
\end{matrix}\right).$$
Performing row operations, we see then that the entries of $C$ can be
reduced modulo $p^{N_{t+1}}$ without changing the lattice. In order
to optimize the size of the objects, we perform this reduction and
define $M_{t+1}$ by:
$$M_{t+1} = \left(\begin{matrix}
M_t & C \text{ mod } p^{N_{t+1}} \smallskip \\ 0 & p^{N_{t+1}}
\end{matrix}\right).$$
We observe in particular that $M_{t+1}$ is still in row-echelon form.

@@

The way the package \ZpLF tracks precision is based on similar 
techniques but differs from \ZpLC in that it does not introduce a cap 
but instead allows $H_t$ to be a sub-$\Zp$-module of $E_t$ of any 
codimension.
This point of view is nice because it implies smaller objects and 
consequently leads to faster algorithms. However, it has a huge 
drawback; indeed, unlike lattices, submodules of $E_t$ of arbitrary 
codimensions are \emph{not} exact objects, in the sense that they cannot 
be represented by integral matrices in full generality. Consequently,
they cannot be encoded on a computer.
We work around this drawback by replacing everywhere exact $p$-adic 
numbers by floating point $p$-adic numbers (at some given precision) 
\cite{course-padic}.

@@

In the \ZpLF framework, $H_{t+1}$ is defined as the image of $H_t$
under the differential $dF_{\underline x}$. Accordingly, the matrix
$M_{t+1}$ is defined as $M_{t+1} = \left(\begin{matrix}
M_t & C \end{matrix}\right)$
where $C$ is the column vector defined by Eq.~\eqref{eq:defC}.
%$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
%\cdot C_i.$$
However, we insist on the fact that all the computations now take
place in the ``ring'' of floating point $p$-adic numbers. Therefore,
we cannot guarantee that the rows of $M_{t+1}$ generate $H_{t+1}$.
Nonetheless, they generate a module which is expected to be close
to $H_{t+1}$.

@@

Unfortunately, checking automatically the assumptions of Proposition 
\ref{prop:precision} in full generality seems to be difficult, though it 
can be done by hand for many particular examples
\cite{caruso-roe-vaccon:14a, caruso:2017, LV16}.

\begin{rmk}
Assuming that Proposition \ref{prop:precision} applies, 
the absolute precision computed as above is optimal if and only if
the inclusion of \eqref{eq:absprec} is an equality. Applying again
Proposition \ref{prop:precision}
with the restricted mapping $\Phi_{t,\ttv} : \calU_t \to \Qp$ and the
lattice $U_t$, we find that this happens if and only if
$d\Phi_{t,\ttv}(U_t) = p^N \Zp$.

@@

In the \ZpLC case, we always have $\cohgt_t(\ttw) = \coind_t(\ttw)$, so 
that the cost becomes $O\big(\coind_t(\ttw)^2\big)$. This does not look 
nice \emph{a priori}. However, the principle of temporal locality 
\cite{denning:2005}
asserts that $\coind_t(\ttw)$ tends to be small in general: destroyed 
variables are often variables that were created recently. As a basic 
example, variables which are local to a small piece of code (\emph{e.g.} 
a short function or a loop) do not survive for a long time. It turns
out that this behavior is typical in many implementations!
%
\begin{figure}
\noindent\hfill%
\begin{tikzpicture}[xscale=0.35,yscale=0.08]
\draw[fill=blue] (-0.45,0) rectangle (0.45,46);
\draw[fill=blue] (0.65,0) rectangle (1.45,69);
\draw[fill=blue] (1.65,0) rectangle (2.45,5);
\draw[fill=blue] (2.65,0) rectangle (3.45,7);
\draw[fill=blue] (3.65,0) rectangle (4.45,7);
\draw[fill=blue] (4.65,0) rectangle (5.45,12);
\draw[fill=blue] (5.65,0) rectangle (6.45,7);
\draw[fill=blue] (6.65,0) rectangle (7.45,5);
\draw[fill=blue] (7.65,0) rectangle (8.45,1);
\draw[fill=blue] (8.65,0) rectangle (9.45,1);
\draw (9.65,0)--(10.45,0);
\draw[fill=blue] (10.65,0) rectangle (11.45,1);
\draw[fill=blue] (11.65,0) rectangle (12.45,1);
\draw[fill=blue] (12.65,0) rectangle (13.45,2);
\draw[fill=blue] (13.65,0) rectangle (14.45,2);
\draw[fill=blue] (14.65,0) rectangle (15.45,2);
\draw[fill=blue] (15.65,0) rectangle (16.45,2);
\draw[fill=blue] (16.65,0) rectangle (17.45,1);
\draw[fill=blue] (17.65,0) rectangle (18.45,1);
\draw[fill=blue] (18.65,0) rectangle (19.45,1);
\node[scale=0.7] at (0,-2.5) { $0$ };
\node[scale=0.7] at (5,-2.5) { $5$ };
\node[scale=0.7] at (10,-2.5) { $10$ };
\node[scale=0.7] at (15,-2.5) { $15$ };
\draw (-2,0)--(-2,72);
\draw (-2.2,0)--(-1.8,0);
\node[scale=0.7,left] at (-2.1,0) { $0$ };
\draw (-2.2,10)--(-1.8,10);
\node[scale=0.7,left] at (-2.1,10) { $10$ };
\draw (-2.2,20)--(-1.8,20);
\node[scale=0.7,left] at (-2.1,20) { $20$ };
\draw (-2.2,30)--(-1.8,30);
\node[scale=0.7,left] at (-2.1,30) { $30$ };
\draw (-2.2,40)--(-1.8,40);
\node[scale=0.7,left] at (-2.1,40) { $40$ };
\draw (-2.2,50)--(-1.8,50);
\node[scale=0.7,left] at (-2.1,50) { $50$ };
\draw (-2.2,60)--(-1.8,60);
\node[scale=0.7,left] at (-2.1,60) { $60$ };
\draw (-2.2,70)--(-1.8,70);
\node[scale=0.7,left] at (-2.1,70) { $70$ };
\end{tikzpicture}%
\hfill\null

>>>>

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
\newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
\newcommand{\QpCR}{\text{\color{output} \rm \tt QpCR}\xspace}
\newcommand{\QpLC}{\text{\color{output} \rm \tt QpLC}\xspace}
\newcommand{\QpLF}{\text{\color{output} \rm \tt QpLF}\xspace}
\newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}

@@

\newcommand{\cIn}{{\color{input} \tt \phantom{Zp}In}:}
\newcommand{\cOut}{{\color{red} \tt \phantom{Z}Out}:}
\newcommand{\cZpCR}{{\color{output} \tt ZpCR}:}
\newcommand{\cZpFP}{{\color{output} \tt ZpFP}:}
\newcommand{\cZpLC}{{\color{output} \tt ZpLC}:}
\newcommand{\cZpLF}{{\color{output} \tt ZpLF}:}
\newcommand{\cQpCR}{{\color{output} \tt QpCR}:}
\newcommand{\cQpLC}{{\color{output} \tt QpLC}:}

@@

If provably correct results are desired, interval arithmetic provides
an alternative to floating point.  Instead of just tracking an approximation to the answer,
one also tracks a radius within which the true result lies.
This method is commonly used for $p$-adic computations since
the ultrametric property of $p$-adic fields frequently keeps the radius small.
Computations remain fairly efficient with this approach, but numerical instability
can still lead to dramatic losses in precision (see \S \ref{sec:demo} for many examples).
Tracking the precision of multiple variables concurrently, the set of possible
true values associated to an inexact value takes the form of an ellipsoid
with axes parallel to the coordinate axes.

@@

In this paper, we describe an implementation of this idea in \sage \cite{sage}.
Rather than attaching a precision to each element, we store the precision of
many elements together by tracking a precision module\footnote{Throughout
the paper, the word ``module'' has its mathematical meaning: it denotes 
an additive group endowed with an external product by scalars in $\Zp$.} 
for the whole collection of variables.  As variables are created and 
destroyed, we update a matrix whose rows represent the vectors in the 
module.  Information about the precision of elements is extracted from 
the matrix as necessary.
Concretely, our package provides two new \sage 
parents\footnote{Parents in \sage are the analogues of container objects 
in computer science. Basically it corresponds to the implementation of a 
concrete mathematical structure, as the ring of $p$-adic integers.}, 
namely \ZpLC and \ZpLF (and their fraction fields \QpLC and \QpLF). The 
acronyms \LC and \LF stand for ``lattice-cap'' and ``lattice-float''
respectively (see \S \ref{sec:implementation} for details). Our package 
has been included in the standard distribution of \sage since version 8.2.

@@

\noindent
Again \ZpCR does not output the optimal precision when the computation
is split into several steps whereas \ZpL does.
These basic examples illustrate situations 
that occur during the execution of many algorithms. 
As a result, interval arithmetic frequently overestimates the loss of precision. 
The aim of our package is to reduce this precision loss.
In the next subsections, we present a bunch of examples showing the
benefit of \ZpL in various contexts.

@@

Many generic algorithms of linear algebra exhibit
instability when used with $p$-adic numbers.
Our package \ZpL frequently eliminates this instability without having to change either the 
algorithm or the implementation.

@@

\noindent \textbf{Characteristic polynomials.}
%
Characteristic polynomials are notoriously hard to compute 
\cite{caruso-roe-vaccon:15, caruso-roe-vaccon:17}.  By default,
\sage uses a division-free algorithm over $\Qp$, this choice is not enough to achieve optimal
precision in the \QpCR case since negative valuations may appear.  We illustrate the different
precision behavior in the following example.

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
 & \verb?M.?{\color{method}\verb?determinant?}\verb?()? \\
\cQpCR
 & \verb?...010000010? \\
\cQpLC
 & \verb?...010000010? \\
\cIn
 & \verb?M.?{\color{method}\verb?charpoly?}\verb?()? \\
\cQpCR
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
\cQpLC
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
 & \verb?...010000010? \\
\end{tabular}}

@@

{\noindent \small
\begin{tabular}{@{}rl}
\cIn
% & {\color{ring}\verb?Q2?}\verb? = Qp(2, prec=50, print_mode="digits")? \\
 & \verb?R.<x,y,z> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, order=?{\color{string}\verb?'invlex'?}\verb?)? \\
 & \verb?F = [ ?{\color{ring}\verb?Q2?}\verb?(2,10)*x + ?{\color{ring}\verb?Q2?}\verb?(1,10)*z, ? \\
 & \verb?      ?{\color{ring}\verb?Q2?}\verb?(1,10)*x^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y^2 - ?{\color{ring}\verb?Q2?}\verb?(2,10)*z^2,? \\
 & \verb?      ?{\color{ring}\verb?Q2?}\verb?(4,10)*y^2 + ?{\color{ring}\verb?Q2?}\verb?(1,10)*y*z + ?{\color{ring}\verb?Q2?}\verb?(8,10)*z^2 ]? \\
\cIn
 & {\color{keyword}\verb?from?}\verb? sage.rings.polynomial.toy_buchberger\?\\
 & \verb?    ?{\color{keyword}\verb?import?}\verb? ?{\color{function}\verb?buchberger_improved?} \\
 & \verb?g = ?{\color{function}\verb?buchberger_improved?}\verb?(ideal(F))? \\
 & \verb?g.?{\color{method}\verb?sort?}\verb?(); g? \\
\cQpCR
 & \verb?[x^3, x*y + ...1100010*x^2,? \\
 & \verb? y^2 + ...11001*x^2, z + ...0000000010*x]? \\
\cQpLC
 & \verb?[x^3, x*y + ...111100010*x^2,? \\
 & \verb? y^2 + ...1111111001*x^2, z + ...0000000010*x]? \\
\end{tabular}}

@@

{\noindent \small
\begin{tabular}{rl}
\cIn
% & \verb?N = 4? \\
% & {\color{ring}\verb?Z2?}\verb? = Zp(2, prec=50, print_mode="digits")? \\
% & {\color{ring}\verb?Q2?}\verb?=?{\color{ring}\verb?Z2?}\verb?.fraction_field()? \\
 & \verb?S.<t> = ?{\color{constructor}\verb?PowerSeriesRing?}\verb?(?{\color{ring}\verb?Q2?}\verb?, 16)? \\
 & \verb?h = 1 + t + t^3? \\
 & \verb?y = t + t^2 * ?{\color{function}\verb?random_element?}\verb?(S, prec=10)? \\
 & \verb?g = y.derivative() / h(y)? \\
 & \verb?u = ?{\color{function}\verb?Newton_Iteration_Solver?}\verb?(g, h, 4); u[15]? \\
\cQpCR
 & \verb?...1101? \\
\cQpLC
 & \verb?...11011101? \\
\end{tabular}}

@@

Let us first examine the first construction [Computation].
With our conventions, if $t$ is the time just before the execution of 
the instruction we are interested in, the $\ttv_i$'s lie in $\calV_t$ 
while $\ttw$ does not. Moreover $\calV_{t+1} = \calV_t \sqcup \{\ttw\}$, 
so that $E_{t+1} = E_t \oplus \Qp e_\ttw$.
The mapping taking the values of variables at time $t$ to that at 
time $t{+}1$ is:
$$\begin{array}{rcl}
F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n)
\end{array}$$
where $x_i$ is the $\ttv_i$-th coordinate of the vector $\underline x$.
The Jacobian matrix of $F$ at $\underline x$ is easily computed; it 
is the block matrix $J_{\underline x}(F) = \big( \begin{matrix} I & L
\end{matrix} \big)$ where $I$ is the identity matrix of size $\text{Card}
\:\calV_t$ and $L$ is the column vector whose $\ttv$-th entry is
$\frac{\partial f}{\partial \ttv} (\underline x)$ if $\ttv$ is one of 
the $\ttv_i$'s and $0$ otherwise.
Therefore, the image of $H_t$ under $dF_{\underline x}$ is represented 
by the matrix $J_{\underline x}(F) \cdot M_t = \big( \begin{matrix} M_t & C
\end{matrix} \big)$ where $C$ is the column vector:
\begin{equation}
\label{eq:defC}
C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
\cdot C_i
\end{equation}
where $C_i$ is the column vector of $M_t$ corresponding to the
variable $\ttv_i$.
Observe that the matrix $J_{\underline x}(F) \cdot M_t$ is no longer a 
square matrix; it has one extra column. This reflects the fact that 
$\dim E_{t+1} = \dim E_t + 1$. Rephrasing this in a different language, 
the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
$E_{t+1}$ but is included in an hyperplane, violating our definition of
a precision lattice.

The package \ZpLC tackles this issue by introducing a cap, analogous
to adding $O(p^N)$ when creating a $p$-adic number from an integer.
Specifically, we choose an integer $N_{t+1}$ and replace the $\Zp$-submodule
$dF_{\underline x}(H_t)$ with the lattice $H_{t+1} = dF_{\underline x}(H_t) \oplus p^{N_{t+1}} \Zp e_\ttw$.
Alternatively, one may introduce the map:
\begin{equation}
\label{eq:tildeF}
\begin{array}{rcl}
\tilde F: \quad E_t \oplus \Qp & \longrightarrow & E_{t+1} \smallskip \\
\underline x \oplus c & \mapsto & \underline x \:\oplus 
\big(f(x_1, \ldots, x_n) + c\big).
\end{array}
\end{equation}
The lattice $H_{t+1}$ is then the image of $H_t \oplus p^{N_{t+1}}
\Zp$ under $d\tilde F_{(\underline x, \star)}$ for any value of $\star$.
In order to carry out this strategy, one must choose $N_{t+1}$.  We do
so by associating two constants to each ring: a relative cap \textsc{relcap}
and an absolute cap \textsc{abscap}. We then set
\begin{equation}
\label{eq:cap}
N_{t+1} = 
\min\big(\textsc{abscap},\, \textsc{relcap} + v_p(y)\big)
\end{equation}
with $y = f(x_1, \ldots, x_n)$.
Larger caps mean that the precision of the final result is less likely to be affected,
but require more space.  See Remark \ref{rmk:cap_optimality} for more details.

In concrete terms, the lattice $H_{t+1}$ is represented by the block 
matrix:
$$\left(\begin{matrix}
M_t & C \smallskip \\ 0 & p^{N_{t+1}}
\end{matrix}\right).$$
Performing row operations, we see then that the entries of $C$ can be
reduced modulo $p^{N_{t+1}}$ without changing the lattice. In order
save space, we perform this reduction and
define $M_{t+1}$ by:
$$M_{t+1} = \left(\begin{matrix}
M_t & C \text{ mod } p^{N_{t+1}} \smallskip \\ 0 & p^{N_{t+1}}
\end{matrix}\right).$$
We observe in particular that $M_{t+1}$ is still in row-echelon form.

@@

The way the package \ZpLF tracks precision is based on similar 
techniques but differs from \ZpLC in that it does not introduce a cap 
but instead allows $H_t$ to be a sub-$\Zp$-module of $E_t$ of any 
codimension.
This point of view is nice because it implies smaller objects and 
consequently leads to faster algorithms. However, it has a huge 
drawback; indeed, unlike lattices, submodules of $E_t$ of arbitrary 
codimensions are \emph{not} exact objects, in the sense that they cannot 
be represented by integral matrices in full generality. Consequently,
they cannot be encoded on a computer.
We work around this drawback by replacing everywhere exact $p$-adic 
numbers by floating point $p$-adic numbers (at some given relative precision $k$) 
\cite{course-padic}.  In the floating point model, we approximate all
numbers in the disc $p^v(u + O(p^k))$ by a single rational number
$p^vu$.  Operations on these numbers are then not exact, but instead
produce the representative in the correct disc.

@@

In the \ZpLF framework, $H_{t+1}$ is defined as the image of $H_t$
under the differential $dF_{\underline x}$. Accordingly, the matrix
$M_{t+1}$ is defined as $M_{t+1} = \left(\begin{matrix}
M_t & C \end{matrix}\right)$
where $C$ is the column vector defined by Eq.~\eqref{eq:defC}.
%$$C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
%\cdot C_i.$$
However, since operations defining $M_{t+1}$ are performed with floating point numbers,
the matrix $M_{t+1}$ gives only an approximate basis for $H_{t+1}$.

@@

Unfortunately, automatically checking the assumptions of Proposition 
\ref{prop:precision} in full generality is difficult because it requires bounds
on higher derivatives.  To produce provably correct results, the assumptions
can be checked by hand \cite{caruso-roe-vaccon:14a, caruso:2017, LV16},
or \ZpCR may be used.

\begin{rmk} \label{rmk:cap_optimality}
Assuming that Proposition \ref{prop:precision} applies, 
the absolute precision computed as above is optimal if and only if
the inclusion of \eqref{eq:absprec} is an equality. Applying again
Proposition \ref{prop:precision}
with the restricted mapping $\Phi_{t,\ttv} : \calU_t \to \Qp$ and the
lattice $U_t$, we find that this happens if and only if
$d\Phi_{t,\ttv}(U_t) = p^N \Zp$.

@@

In the \ZpLC case, we always have $\cohgt_t(\ttw) = \coind_t(\ttw)$, so 
that the cost becomes $O\big(\coind_t(\ttw)^2\big)$, which seems high \emph{a priori}.
However, the principle of temporal locality \cite{denning:2005}
asserts that $\coind_t(\ttw)$ tends to be small: destroyed 
variables are often recently created ones.  As a simple 
example, variables which are local to a small piece of code (\emph{e.g.} 
a short function or a loop) have a short lifetime.
%
\begin{figure}
\noindent\hfill%
\begin{tikzpicture}[xscale=0.35,yscale=0.08]
\draw[fill=blue] (-0.45,0) rectangle (0.45,46);
\draw[fill=blue] (0.65,0) rectangle (1.45,69);
\draw[fill=blue] (1.65,0) rectangle (2.45,5);
\draw[fill=blue] (2.65,0) rectangle (3.45,7);
\draw[fill=blue] (3.65,0) rectangle (4.45,7);
\draw[fill=blue] (4.65,0) rectangle (5.45,12);
\draw[fill=blue] (5.65,0) rectangle (6.45,7);
\draw[fill=blue] (6.65,0) rectangle (7.45,5);
\draw[fill=blue] (7.65,0) rectangle (8.45,1);
\draw[fill=blue] (8.65,0) rectangle (9.45,1);
\draw (9.65,0)--(10.45,0);
\draw[fill=blue] (10.65,0) rectangle (11.45,1);
\draw[fill=blue] (11.65,0) rectangle (12.45,1);
\draw[fill=blue] (12.65,0) rectangle (13.45,2);
\draw[fill=blue] (13.65,0) rectangle (14.45,2);
\draw[fill=blue] (14.65,0) rectangle (15.45,2);
\draw[fill=blue] (15.65,0) rectangle (16.45,2);
\draw[fill=blue] (16.65,0) rectangle (17.45,1);
\draw[fill=blue] (17.65,0) rectangle (18.45,1);
\draw[fill=blue] (18.65,0) rectangle (19.45,1);
\node[scale=0.7] at (0,-2.5) { $0$ };
\node[scale=0.7] at (5,-2.5) { $5$ };
\node[scale=0.7] at (10,-2.5) { $10$ };
\node[scale=0.7] at (15,-2.5) { $15$ };
\draw (-2,0)--(-2,72);
\draw (-2.2,0)--(-1.8,0);
\node[scale=0.7,left] at (-2.1,0) { $0$ };
\draw (-2.2,10)--(-1.8,10);
\node[scale=0.7,left] at (-2.1,10) { $10$ };
\draw (-2.2,20)--(-1.8,20);
\node[scale=0.7,left] at (-2.1,20) { $20$ };
\draw (-2.2,30)--(-1.8,30);
\node[scale=0.7,left] at (-2.1,30) { $30$ };
\draw (-2.2,40)--(-1.8,40);
\node[scale=0.7,left] at (-2.1,40) { $40$ };
\draw (-2.2,50)--(-1.8,50);
\node[scale=0.7,left] at (-2.1,50) { $50$ };
\draw (-2.2,60)--(-1.8,60);
\node[scale=0.7,left] at (-2.1,60) { $60$ };
\draw (-2.2,70)--(-1.8,70);
\node[scale=0.7,left] at (-2.1,70) { $70$ };
\end{tikzpicture}%
\hfill\null
